### 2025-06-09 lecturenotes_aima203/ch4_aima203.tex
interpolation

\setcounter{chapter}{3}
\begin{document}
\chapter{Interpolation}
\section{Introduction}
The central problem of numerical analysis is: given a set of tabular values $(x_0,y_0), (x_1,y_1),(x_2,y_2),...,(x_n,y_n)$ satisfying the realtion $y=f(x)$, where the explict nature of $f(x)$ is not known, is to find a simpler function $\phi (x)$ such that $\phi (x)$ and $f(x)$ agree at the set of the tabulated points. Such process is called \textbf{interpolation}. If $\phi (x)$ is a polynomial, then the process is called \textit{polynomial interpolation}, and $\phi (x)$ a interpolating polynomial. In this chapter we are concerned with only polynomial interpolation.

@@

\section{Interpolation with Unevenly spaced points}
This section discusses the inerpolation method for unequally spaced values of the argument.

@@

\subsection{Disadvantage of Lagrange's Interpolation}
The Lagrange's Interpolation formula has the disadvantage that if another interpolation point were added, then interpolation coefficients $l_i(x)$ will have to be recomputed. We therefore see an interpolation polynomial which has the property that a polynomial of higer degree may be derived from it by simply adding new terms. Newton's general interpolation formula is one such formula.

\section{Divided Differences}
Given $(n+1)$ points, $(x_0,y_0), \, (x_1,y_1), \, ..., \, (x_n,y_n)$ of $x$ and $y(x)$, with $x$ values equally spaced, the divided differences of order $1,2,...,n$ are defined by the relations:
\begin{align*}
  [x_0,x_1] &= \frac{y_1 - y_0}{x_1 - x_0} \\[1mm]
  [x_0,x_1,x_2] &= \frac{[x_1, x_2]-[x_0,x_1]}{x_2 - x_0} \\[1mm]
  ... \\
  [x_0,x_1,x_2,...,x_n] &= \frac{[x_1, x_2,...,x_n]-[x_0,x_1,...,x_{n-1}]}{x_n - x_0}
\end{align*}
It is easy to see that, $\displaystyle [x_0,x_1] = \frac{y_1 - y_0}{x_1 - x_0} = \frac{y_0}{x_0 - x_1} + \frac{y_1}{x_1 - x_0} = [x_1,x_0]$.
Again,
\begin{align*}
  [x_0,x_1,x_2] &= \frac{[x_2, x_0]-[x_0,x_1]}{x_2 - x_0} \\[1mm]
                &= \frac{1}{x_2 - x_0} \left [ \frac{y_2- y_1}{x_2 - x_1} - \frac{y_1 - y_0}{x_1 - x_0} \right ] \\[1mm]
                  &= \frac{1}{x_2 - x0} \left [ \frac{y_2}{x_2 - x_1} -y_1 \left ( \frac{1}{x_2-x_1} + \frac{1}{x_1 - x_0} \right ) + \frac{y_0}{x_1 - x_0} \right ] \\[1mm]
  &= \frac{y_0}{(x_0-x_1)(x_0-x_2)} + \frac{y_1}{(x_1-x_0)(x_1-x_2)} + \frac{y_2}{(x_2-x_0)(x_2-x_1)}
\end{align*}

\subsection{Newton's General Interpolation Formula}
We have,  $\displaystyle [x,x_0] = \frac{y - y_0}{x - x_0} \implies y = y_0 + (x-x_0)[x,x_0]$ \\[1mm]
And $\displaystyle [x,x_0,x_1] =  \frac{[x, x_0]-[x_0,x_1]}{x - x_1} \implies = [x,x_0] = [x_0,x_1] + (x-x_1)[x,x_0,x_1]$ \\[1mm]
From these two relations we get,
\begin{equation}
  \label{ninter}
  y = y_0 + (x-x_0)[x_0,x_1] + (x-x_0)(x-x_1)[x,x_0,x_1]
\end{equation}
Similarly we can get $[x,x_0,x_1] =  [x_0,x_1,x_2] + (x-x_2)[x,x_0,x_1,x_2]$ and so on...
Proceeding this way we get,
\begin{equation}
  \label{eq:ngeneral}
  \begin{gathered}
    y = y_0 + (x-x_0)[x_0,x_1] + (x-x_0)(x-x_1)[x_0,x_1,x_2] + \\[1mm]
    (x-x_0)(x-x_1)(x-x_2)[x_0,x_1,x_2,x_3] + ... \\[1mm]
    + (x-x_0)(x-x_1)(x-x_2)...(x-x_n)[x,x_0,x_1,...,x_n]
  \end{gathered}
  \end{equation}
  This Equation~\ref{eq:ngeneral} is called the Newton's general interpolation formula with divided differnces.

>>>>

\setcounter{chapter}{3}
\begin{document}
\chapter{Interpolation}
\section{Introduction}
The central problem of numerical analysis is: given a set of tabular values \\
$(x_0,y_0), (x_1,y_1),(x_2,y_2),...,(x_n,y_n)$ satisfying the realtion $y=f(x)$, where the explict nature of $f(x)$ is not known, is to find a simpler function $\phi (x)$ such that $\phi (x)$ and $f(x)$ agree at the set of the tabulated points. Such process is called \textbf{interpolation}. If $\phi (x)$ is a polynomial, then the process is called \textit{polynomial interpolation}, and $\phi (x)$ a interpolating polynomial. In this chapter we are concerned with only polynomial interpolation.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=11cm]{newt_forward.png}
\end{figure}

\begin{enumerate}
\item Find the cubic polynomial which takes the following values: $y(1)=24, \;\;\;  y(3)=120,  \;\;\; y(5)=336,  \;\;\; y(7)=720$. Then obtain the value of $y(8)$.
\item Find a cubic polynomial which fits the data: $(-2, -12)  \;\;\; (-1, -8),  \;\;\; (2,3),  \;\;\; (3,5)$.
\end{enumerate}
\section{Interpolation with Unevenly spaced points}
This section discusses the inerpolation method for unequally spaced values of the argument.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=12cm]{lagra_ex.png}
\end{figure}
\subsection{Disadvantage of Lagrange's Interpolation}
The Lagrange's Interpolation formula has the disadvantage that if another interpolation point were added, then interpolation coefficients $l_i(x)$ will have to be recomputed. We therefore see an interpolation polynomial which has the property that a polynomial of higer degree may be derived from it by simply adding new terms. Newton's general interpolation formula is one such formula.

\section{Divided Differences}
Given $(n+1)$ points, $(x_0,y_0), \, (x_1,y_1), \, ..., \, (x_n,y_n)$ of $x$ and $y(x)$, with $x$ values equally spaced, the divided differences of order $1,2,...,n$ are defined by the relations:
\begin{align*}
  [x_0,x_1] &= \frac{y_1 - y_0}{x_1 - x_0} \\[1mm]
  [x_0,x_1,x_2] &= \frac{[x_1, x_2]-[x_0,x_1]}{x_2 - x_0} \\[1mm]
  ... \\
  [x_0,x_1,x_2,...,x_n] &= \frac{[x_1, x_2,...,x_n]-[x_0,x_1,...,x_{n-1}]}{x_n - x_0}
\end{align*}
It is easy to see that, $\displaystyle [x_0,x_1] = \frac{y_1 - y_0}{x_1 - x_0} = \frac{y_0}{x_0 - x_1} + \frac{y_1}{x_1 - x_0} = [x_1,x_0]$.
Again,
\begin{align*}
  [x_0,x_1,x_2] &= \frac{[x_2, x_0]-[x_0,x_1]}{x_2 - x_0} \\[1mm]
                &= \frac{1}{x_2 - x_0} \left [ \frac{y_2- y_1}{x_2 - x_1} - \frac{y_1 - y_0}{x_1 - x_0} \right ] \\[1mm]
                  &= \frac{1}{x_2 - x_0} \left [ \frac{y_2}{x_2 - x_1} -y_1 \left ( \frac{1}{x_2-x_1} + \frac{1}{x_1 - x_0} \right ) + \frac{y_0}{x_1 - x_0} \right ] \\[1mm]
  &= \frac{y_0}{(x_0-x_1)(x_0-x_2)} + \frac{y_1}{(x_1-x_0)(x_1-x_2)} + \frac{y_2}{(x_2-x_0)(x_2-x_1)}
\end{align*}
\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=5cm]{divided_for.png}
\end{figure}
\subsection{Newton's General Interpolation Formula}
We have,  $\displaystyle [x,x_0] = \frac{y - y_0}{x - x_0} \implies y = y_0 + (x-x_0)[x,x_0]$ \\[1mm]
And $\displaystyle [x,x_0,x_1] =  \frac{[x, x_0]-[x_0,x_1]}{x - x_1} \implies = [x,x_0] = [x_0,x_1] + (x-x_1)[x,x_0,x_1]$ \\[1mm]
From these two relations we get,
\begin{equation}
  \label{ninter}
  y = y_0 + (x-x_0)[x_0,x_1] + (x-x_0)(x-x_1)[x,x_0,x_1]
\end{equation}
Similarly we can get $[x,x_0,x_1] =  [x_0,x_1,x_2] + (x-x_2)[x,x_0,x_1,x_2]$ and so on...
Proceeding this way we get,
\begin{equation}
  \label{eq:ngeneral}
  \begin{gathered}
    y = y_0 + (x-x_0)[x_0,x_1] + (x-x_0)(x-x_1)[x_0,x_1,x_2] + \\[1mm]
    (x-x_0)(x-x_1)(x-x_2)[x_0,x_1,x_2,x_3] + ... \\[1mm]
    + (x-x_0)(x-x_1)(x-x_2)...(x-x_n)[x,x_0,x_1,...,x_n]
  \end{gathered}
  \end{equation}
  This Equation~\ref{eq:ngeneral} is called the Newton's general interpolation formula with divided differences.

  \begin{figure}[h!]
  \centering
\includegraphics[width=13cm, height=11cm]{divided.png}
\end{figure}

### 2025-06-09 question_papers/aima203_first_in_sem.tex
interpolation

\documentclass[12pt]{exam}
\usepackage{preamble_question}
\usepackage[normalem]{ulem}
\renewcommand\ULthickness{2pt}
\setlength\ULdepth{1.3ex}

@@

\begin{center}
 {\bfseries  {\Large Kathmandu University}} \\[1mm]
 First In-Semester Exam-2025\\[-1mm]
 {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
\end{center}
\vskip 5mm

\begin{minipage}{0.70\linewidth}
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: II \\
    Time: 60 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 203 \\
    Semester: II \\
    F.M. : 20
  \end{flushleft}
\end{minipage}
\vskip 2mm
\rule{\textwidth}{1pt}
\hskip 15mm \textit{Write the answer according to the mark allocated to it}. \textit{For short questions, write short answers, for long questions write long answers}.

\vskip 10mm

@@

\question

\end{questions}
\end{document}

>>>>

\documentclass[12pt]{exam}
\usepackage{preamble_question_small}
\usepackage[normalem]{ulem}
\renewcommand\ULthickness{2pt}
\setlength\ULdepth{1.3ex}

@@

\begin{center}
 {\bfseries  { Kathmandu University}} \\[-1mm]
 First In-Semester Exam-2025\\[-2mm]
 {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
\end{center}
\begin{minipage}{0.70\linewidth}
  \small
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: II \\
    Time: 60 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 203 \\
    Semester: II \\
    F.M. : 20
  \end{flushleft}
\end{minipage}
\vskip 1mm
\rule{\textwidth}{1pt}
\vskip -5mm
\begin{questions}

\question Find a root of $x^3-4x-9=0$, using bisection method, with absolute error less than $10^2$. [4]

\question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]
\end{questions}

\vspace{3cm}
\begin{center}
 {\bfseries  { Kathmandu University}} \\[-1mm]
 First In-Semester Exam-2025\\[-2mm]
 {\small \textbf{Department of Artificial Intelligence, Panchkhal}}
\end{center}
\begin{minipage}{0.70\linewidth}
  \small
  \begin{flushleft}
    Level: B.Tech Artificial Intelligence \\
    Year: II \\
    Time: 60 minutes
  \end{flushleft}
\end{minipage} \hfill
\begin{minipage}{0.25\linewidth}
  \begin{flushleft}
    Course: AIMA 203 \\
    Semester: II \\
    F.M. : 20
  \end{flushleft}
\end{minipage}
\vskip 1mm
\rule{\textwidth}{1pt}

@@

\question Find a root of $x^3-4x-9=0$, using bisection method, with absolute error less than $10^2$. [4]

\question Describe error due to truncation. Find a root of \(f(x)=xe^x-1=0\), correct to \(2\) decimal places. [1+3]

  \clearpage

\vspace{4cm}
\question Find a real root of the equation $x^2+y =1, x+y^2=7$, using the method of iteration, correct up-to 2 decimal places. [4]

\question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
\vskip -5mm
\question Show that the matrix $ \small A=   \begin{bmatrix}
    25  & 24 & 10 \\
    66 & 78 & 37 \\
    92 & -73 & -80
  \end{bmatrix}
  $ is ill-conditioned.
\vspace{12cm}

\question Find a real root of the equation $x^2+y =1, x+y^2=7$, using the method of iteration, correct up-to 2 decimal places. [4]

\question Solve using LU-factorization method. $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$. [4]
\vskip -5mm
\question Show that the matrix $ \small A=   \begin{bmatrix}
    25  & 24 & 10 \\
    66 & 78 & 37 \\
    92 & -73 & -80
  \end{bmatrix}
  $ is ill-conditioned.
\end{questions}
\end{document}
